Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Some changes to Section5: minor redrafts; more information around the DWARF verification

At the bottom of the stack, we have the |handler_info|, which contains the
pointer to parent fiber, and the closures for the value, exception and effect
cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a block to
support DWARF unwinding, whose details we will return to later. Then, there is
a top-level handler that invokes the exception handler closure |clos_hexn|
after switching to the parent fiber. This is followed by the |pc| of the code
that invokes value handler |clos_hval| after switching to the parent stack. The
stack is laid out such that when the computation in the handler returns,
control goes to value handler.

We have the variable-sized area for the OCaml frames next. Since we want the
handlers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| goes less than stack threshold (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
We observed that most function calls are to leaf functions with small frame
size. In order to take advantage of this fact, we introduce a small red zone at
the top of the stack. The compiler elides the stack overflow check for leaf
functions whose frame size is less than the size of the red zone.

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
observed that directly using |mimalloc|~\cite{Leijen19} outperforms the stack
cache.

@@

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the exception and stack pointers in the current fiber, saving the
allocation pointer in the |Caml_state|, updating |rsp| to the top of system
stack (maintained in |Caml_state|), and calling the C function. The actions are
reversed when returning from the external call. One subtlety is that for C
functions that take arguments on the stack, the arguments must be copied to the
C stack from the OCaml stack where they are laid out. For vast majority of
external calls, we perform only a few additional instructions over stock OCaml.

@@

Logically, DWARF maintains a large table which records for every machine
instruction where the return address and callee-saved registers are stored. To
avoid reifying this large table, DWARF represents the table using a compact,
turing-complete, bytecode representation, which is interpreted on demand for
building the stack unwind table. Within each function, DWARF maintains a
\emph{canonical frame address} (CFA), which is invariant in the function body,
and is traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which we have validated with the DWARF
unwind table validator tool~\cite{Bastian19}. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callback, we
save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented similarly by saving the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is neither continued nor discontinued. Since fibers are
malloced and are |free|d when the computation being handled returns a value
or raises an exception, not resuming continuations leaks memory. In addition,
unresumed continuations may also leak other resource such as open file
descriptors.

>>>>

At the bottom of the stack, we have the |handler_info|, which contains the
pointer to parent fiber, and the closures for the value, exception and effect
cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a block to
support DWARF unwinding, whose details we will return to later. Then, there is
a top-level handler that invokes the exception handler |clos_hexn|
after switching to the parent fiber. This is followed by the |pc| of the code
that invokes the value handler |clos_hval| after switching to the parent stack. The
stack is laid out such that when the computation in the handler returns,
control goes to the value handler.

We have the variable-sized area for the OCaml frames next. Since we want the
handlers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| becomes less than the stack threshold (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
We observed that most function calls are to leaf functions with small frame
size. In order to take advantage of this fact, we introduce a small red zone at
the top of the stack. The compiler elides the stack overflow check for leaf
functions whose frame size is less than the size of the red zone.

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
observed that directly using |mimalloc|~\cite{Leijen19} outperforms a stack
cache.

@@

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the stack pointer in the current fiber, saving the
allocation pointer in the |Caml_state|, updating |rsp| to the top of system
stack (maintained in |Caml_state|), and calling the C function. The actions are
reversed when returning from the external call. For C
functions that take arguments on the stack, the arguments must be copied to the
C stack from the OCaml stack where they are laid out. In the vast majority of
external calls, we perform only a few additional instructions over stock OCaml.

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. This representation
is interpreted on demand for building the stack unwind table. Within each
function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on x86-64,
where the return address is pushed on the stack on call, the return address is
at |CFA - 8|.

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers. We have utilized a DWARF verification tool~\cite{Bastian19} with
extensions to handle PIE and additional DWARF primitives to flag possible errors
in the hand written DWARF directives where we have altered the OCaml assembler
to implement effects. In three places additional manual verification was
required; two were where the stack is intentionally switched between fibers and
one was where the return to the parent stack on child stack completion could not
track stack resizes and intermediate effect handler frames.

@@

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is neither continued nor discontinued. Since fibers are
|malloc|ed and are |free|d when the computation being handled returns a value
or raises an exception, not resuming continuations leaks memory. In addition,
unresumed continuations may also leak other resource such as open file
descriptors.
